---
title: "On Rectangular Map Approximations"
author: "Christian Panse"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{recmap primer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Intro
This package contains an implementation of the recmap algorithm^[Heilmann R, 
Keim D, Panse C and Sips M (2004). “RecMap: Rectangular Map Approximations.” 
In _InfoVis 2004, IEEE Symposium on
Information Visualization, Austin, Texas_, pp. 33-40.
[DOI:10.1109/INFVIS.2004.57](http://dx.doi.org/10.1109/INFVIS.2004.57)].
The input contains for each map region 

- a x-y coordinate, 
- a map ratio, 
- and a statistical value.

The x-y coordinates representing the centers of gravity, the ratio the lat/long 
ratio, and the  statistical values define the desired 
area of each map region. 
The output is a rectangular cartogram where the map regions are 

- non overlapping, 
- connected, 
- ratio and area of each rectangle corresponde to the desired areas.

The construction heuristing places the map region by preserving important 
geospatial constraints, in particular

- the topology of the pseudo dual,
- the relative position, 
- and the ratios of each input map regions.

```{r}
library(recmap)
```
```{r echo=FALSE}
.draw_recmap_us_state_ev <- function(...){

  cm <- c("#FF0000", "#FF0505", "#FF0A0A", "#FF1010", "#FF1515", "#FF1A1A", "#FF1F1F",
    "#FF2424", "#FF2A2A", "#FF2F2F", "#FF3434", "#FF3939", "#FF3E3E", "#FF4444",
    "#FF4949", "#FF4E4E", "#FF5353", "#FF5858", "#FF5E5E", "#FF6363", "#FF6868",
    "#FF6D6D", "#FF7272", "#FF7878", "#FF7D7D", "#FF8282", "#FF8787", "#FF8D8D",
    "#FF9292", "#FF9797", "#FF9C9C", "#FFA1A1", "#FFA7A7", "#FFACAC", "#FFB1B1",
    "#FFB6B6", "#FFBBBB", "#FFC1C1", "#FFC6C6", "#FFCBCB", "#FFD0D0", "#FFD5D5",
    "#FFDBDB", "#FFE0E0", "#FFE5E5", "#FFEAEA", "#FFEFEF", "#FFF5F5", "#FFFAFA",
    "#FFFFFF", "#FFFFFF", "#FAFAFF", "#F5F5FF", "#EFEFFF", "#EAEAFF", "#E5E5FF",
    "#E0E0FF", "#DBDBFF", "#D5D5FF", "#D0D0FF", "#CBCBFF", "#C6C6FF", "#C1C1FF",
    "#BBBBFF", "#B6B6FF", "#B1B1FF", "#ACACFF", "#A7A7FF", "#A1A1FF", "#9C9CFF",
    "#9797FF", "#9292FF", "#8D8DFF", "#8787FF", "#8282FF", "#7D7DFF", "#7878FF",
    "#7272FF", "#6D6DFF", "#6868FF", "#6363FF", "#5E5EFF", "#5858FF", "#5353FF",
    "#4E4EFF", "#4949FF", "#4444FF", "#3E3EFF", "#3939FF", "#3434FF", "#2F2FFF",
    "#2A2AFF", "#2424FF", "#1F1FFF", "#1A1AFF", "#1515FF", "#1010FF", "#0A0AFF",
    "#0505FF", "#0000FF")
  
  recmap_us_state_ev.file <- system.file("extdata", 
                                         "recmap_us_state_ev.polygon", 
                                         package = "recmap")
  
  recmap_us_state_ev <- read.table(recmap_us_state_ev.file, sep = '|', 
                                   col.names=c('x', 'y'))
  
  us_state_election_2004.file <- system.file("extdata", 
                                         "us_state_election_2004.csv", 
                                         package = "recmap")
  
  us_state_election_2004 <- read.table(us_state_election_2004.file, 
                                       sep = ',')
  
  plot(recmap_us_state_ev$x, recmap_us_state_ev$y,type='n', 
       axes = FALSE, 
       asp = 1,
       xlab='', 
       ylab='')
  
  polygon(recmap_us_state_ev$x, recmap_us_state_ev$y,
          col=cm[round(length(cm)*(us_state_election_2004$V8/(us_state_election_2004$V8 + us_state_election_2004$V9)))+1])
  
  text(us_state_election_2004$V1,us_state_election_2004$V2-7,
       as.character(us_state_election_2004$V3),cex=round(us_state_election_2004$V5*20)/100,
       lwd=2.5,
       pos=3,
       col="black");
}
```
```{r fig.width=7, fig.retina=2, fig.cap="Rectangular Cartogram of the U.S. election 2004; area corresponds to the number of electors."}
op<-par(mar=c(0,0,0,0), bg='black')
.draw_recmap_us_state_ev()
```

